---
title: "BC population and geography data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BC population and geography data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`islhepi` provides a controlled access layer around the BC Data Catalogue.
Permanent record and resource identifiers are pinned in the package, downloaded
objects use consistent join keys, and no request is made when the package loads.

## Population denominators

The population helper supports Local Health Areas and Health Service Delivery
Areas. It reshapes one-year ages into the same age bands used by
`islh_age_group()`.

```{r, eval=FALSE}
library(dplyr)
library(islhepi)

lha_population <- islh_bc_population(
  geography = "lha",
  years = 2021:2026,
  sex = c("F", "M"),
  age_breaks = "five_year"
)
```

Female and male rows are returned separately. Summing them produces a total
without double counting. Alternatively, request `sex = "T"` on its own when
the catalogue total is required.

## Health-geography boundaries

The geography helper supports health authorities, HSDAs, LHAs and CHSAs. It
returns an `sf` object in BC Albers by default and limits the result to Island
Health.

```{r, eval=FALSE}
lha_boundaries <- islh_bc_geography("lha")

# Use all provincial boundaries when needed.
bc_chsa <- islh_bc_geography("chsa", health_authority = NULL)
```

`health_authority = "Island Health"` and `"Vancouver Island"` select the same
catalogue records.

## A checked population join

Use the standardized geography type and code for a transparent join. A strict
one-to-one relationship makes duplicate or unmatched areas fail before a map
is drawn.

```{r, eval=FALSE}
lha_extract <- islh_bc_population("lha", years = 2025, sex = "T")
lha_source <- attr(lha_extract, "bcdata_source")

lha_total <- lha_extract |>
  summarise(
    population = sum(population),
    .by = c(geography, geography_code, year, estimate_type)
  )

lha_map_data <- inner_join(
  islh_bc_geography("lha"),
  lha_total,
  by = join_by(geography, geography_code),
  relationship = "one-to-one",
  unmatched = "error",
  na_matches = "never"
)
```

Read the `bcdata_source` attribute immediately after download, or use
`islh_bc_sources()`, to record catalogue provenance. Attributes may be dropped
by later data transformations, so persist source metadata explicitly when it
must travel with an analytical extract.

## Reproducible extracts

The package retrieves current catalogue content and deliberately does not ship
copied datasets. For a governed analysis, save the retrieved object in the
project with its retrieval date and source metadata. Re-run the retrieval only
when the analysis protocol permits the denominator or geography version to
change.

If a request fails on a health-authority network, verify that
`catalogue.data.gov.bc.ca` is reachable through the configured proxy. The
package does not modify `bcdata` internals or bypass its connectivity checks.
