---
title: "Getting started with islhepi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with islhepi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`islhepi` provides small, validated functions for routine public health
analysis. It is separate from `islhr`, which handles Island Health branding,
tables and report scaffolding.

## Rates

```{r}
library(islhepi)

islh_crude_rate(cases = 12, population = 50000)
```

`islh_ci_poisson()` provides Byar and exact intervals. Byar is the default for
continuity with existing analyses. Use the exact method for small counts when
required by the governing analytical standard.

```{r}
islh_ci_poisson(c(0, 3, 25), method = "exact")
```

For age-standardised rates, pass one value per stratum and the chosen standard
population. The function refuses missing strata because they would make the
single summary rate incomplete.

```{r}
cases <- c(5, 12, 40, 80)
population <- c(20000, 25000, 22000, 15000)
standard <- c(30000, 30000, 25000, 15000)

islh_dsr(cases, population, standard)
```

## Disclosure control

There is deliberately no default suppression threshold or rounding base.
Those are release-policy decisions and must be supplied explicitly.

```{r}
islh_suppress(c(0, 3, 42), threshold = 5)

counts <- data.frame(
  area = c("North", "Central", "South"),
  cases = c(3, 42, 17)
)
islh_suppress_table(
  counts,
  "cases",
  threshold = 5,
  complementary = TRUE
)
```

Complementary suppression in this helper operates independently within each
selected column. It does not establish protection for multidimensional tables,
linked releases or tables with several margins. Review those outputs against
the applicable disclosure-control policy before release.

