---
title: "Getting started with islhepi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with islhepi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(islhepi)
```

`islhepi` provides small, validated functions for routine public health
analysis. Every example on this page is evaluated when the website is built,
so the printed values are actual package output.

The package implements calculation mechanics. It does not choose the
suppression threshold, standard population or interval method for a release.

## Choose the function

| Task | Function | Output |
|---|---|---|
| Group individual ages | `islh_age_group()` | Ordered factor |
| Calculate a crude rate | `islh_crude_rate()` | Data frame with rate and interval |
| Calculate a Poisson count interval | `islh_ci_poisson()` | Data frame with lower and upper limits |
| Calculate a directly standardised rate | `islh_dsr()` | One-row data frame |
| Hide small counts | `islh_suppress()` | Numeric or labelled character vector |
| Apply suppression to a table | `islh_suppress_table()` | Updated data frame |
| Round counts | `islh_round_base()` | Numeric vector |
| Retrieve BC denominators | `islh_bc_population()` | Tidy population table |
| Retrieve BC boundaries | `islh_bc_geography()` | Standardised `sf` object |

## Group ages

The result is an ordered factor, which preserves the correct order in tables
and plots.

```{r age-groups}
people <- data.frame(age = c(0, 4, 18, 42, 67, 91, NA))
people$age_group <- islh_age_group(people$age, breaks = "broad")
people
```

Built-in choices are `"five_year"`, `"phac"` and `"broad"`. A local set of
lower bounds also works:

```{r custom-age-groups}
islh_age_group(c(2, 17, 42, 91), breaks = c(0, 18, 65))
```

Negative and missing ages become `NA`. Check those rows before calculating a
rate rather than silently dropping them.

## Crude rates

`islh_crude_rate()` returns the numerator and denominator beside the estimate
and confidence limits, making the calculation easy to audit.

```{r crude-rate}
crude <- islh_crude_rate(
  cases = c(12, 45),
  population = c(50000, 120000),
  per = 100000,
  method = "exact"
)
round(crude, 1)
```

Read the first row as: 12 events in a population of 50,000, expressed per
100,000 people, with the chosen Poisson interval. Counts may exceed the
denominator when the numerator contains repeat events and the denominator is
person-time.

## Poisson count intervals

```{r poisson-intervals}
round(islh_ci_poisson(c(0, 3, 25), method = "exact"), 2)
```

Use `method = "exact"` for small counts when required by the analytical
standard. `"byar"` is an approximation and falls back to the exact upper
limit when the count is zero.

## Directly standardised rates

Pass one value per age stratum. All three vectors must be aligned in the same
order.

```{r standardised-rate}
age_specific <- data.frame(
  age_group = c("0-19", "20-44", "45-64", "65+"),
  cases = c(5, 12, 40, 80),
  population = c(20000, 25000, 22000, 15000),
  standard_population = c(30000, 30000, 25000, 15000)
)

dsr <- with(
  age_specific,
  islh_dsr(cases, population, standard_population)
)
round(dsr, 1)
```

The returned `population` is the total observed denominator. The standard
population is converted to weights and is not added to the result.
`method = "gamma"` uses the Fay-Feuer interval; `"normal"` is available for
comparison with analyses that explicitly require it.

## Suppress small cells

There is deliberately no default threshold. Supply the approved rule for the
specific release.

```{r suppress-vector}
counts <- c(0, 3, 5, 17, 42)
islh_suppress(counts, threshold = 5, label = "Suppressed")
```

Zero is kept because it is an observed absence, not a small positive count.
With the default `inclusive = TRUE`, both 3 and 5 are suppressed.

For a table:

```{r suppress-table}
area_counts <- data.frame(
  area = c("North", "Central", "South"),
  cases = c(3, 42, 17),
  contacts = c(1, 55, 4)
)

islh_suppress_table(
  area_counts,
  cols = c("cases", "contacts"),
  threshold = 5,
  label = "Suppressed"
)
```

Complementary suppression can hide a second cell when one hidden value could
be recovered from a visible total:

```{r complementary-suppression}
islh_suppress_table(
  area_counts,
  cols = "cases",
  threshold = 5,
  label = "Suppressed",
  complementary = TRUE
)
```

This helper works independently within each selected column. It does not prove
that a multidimensional or linked release is protected.

## Round counts

```{r deterministic-rounding}
islh_round_base(c(0, 2, 3, 7, 12, 43), base = 5)
```

Deterministic rounding returns the same values each time. Random rounding is
unbiased over repeated use but varies between runs. Set a seed, round once and
save the released values:

```{r random-rounding}
set.seed(42)
islh_round_base(c(2, 3, 7, 12), base = 5, method = "random")
```

## Population and geography data

The BC Data Catalogue helpers require a network request and are documented in
[BC population and geography data](bcdata.html). That article shows the exact
returned columns, representative output, a checked join and troubleshooting
for unmatched rows.

Use `islhr` after the calculation when the result needs an Island Health
figure, table or Quarto report.
