---
title: "Routine surveillance with islhepi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Routine surveillance with islhepi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(islhepi)
```

This guide demonstrates a small, complete workflow for daily or weekly
surveillance. Every chunk is evaluated when the pkgdown site is built, so the
tables below are actual package output.

The package handles reusable mechanics. It does not decide case definitions,
alert policy or the public-health response to a statistical exceedance.

## Example event data

The simulated line list has one row per encounter, a date, a site and a case
classification.

```{r event-data}
set.seed(42)

events <- data.frame(
  encounter_id = sprintf("E%03d", 1:90),
  event_date = as.Date("2026-06-01") + sample(0:41, 90, replace = TRUE),
  site = sample(c("Central", "North", "South"), 90, replace = TRUE),
  classification = sample(
    c("Confirmed", "Probable"),
    90,
    replace = TRUE,
    prob = c(0.8, 0.2)
  )
)

head(events)
```

## Check the records first

`islh_check_events()` returns one row for every problem. It does not alter the
line list or remove affected records.

```{r clean-event-check}
islh_check_events(
  events,
  id = encounter_id,
  date = event_date,
  required = c(site, classification),
  max_date = NULL
)
```

A deliberately damaged example shows the review output:

```{r issue-example}
events_with_issues <- events[1:5, ]
events_with_issues$encounter_id[2] <- events_with_issues$encounter_id[1]
events_with_issues$event_date[3] <- NA
events_with_issues$site[4] <- NA

islh_check_events(
  events_with_issues,
  id = encounter_id,
  date = event_date,
  required = site,
  max_date = NULL
)
```

Use the `.row` values to display the affected source records to the data owner.

## Create complete daily counts

`groups` lists the sites that must appear even when a site has no events during
the reporting window. `fill = TRUE` also creates days when nobody had an event.

```{r daily-counts}
daily <- islh_count_events(
  events,
  date = event_date,
  id = encounter_id,
  by = site,
  interval = "day",
  from = "2026-06-01",
  to = "2026-07-12",
  fill = TRUE,
  groups = c("Central", "North", "South")
)

head(daily, 9)
```

For CDC epidemiological weeks, use `interval = "epiweek"`. ISO weeks use
`"isoweek"`. For a locally anchored week, use `"week"` and provide
`week_start`, where 1 is Monday and 7 is Sunday.

```{r weekly-counts}
weekly <- islh_count_events(
  daily,
  date = period_start,
  count = count,
  by = site,
  interval = "week",
  week_start = 1,
  from = "2026-06-01",
  to = "2026-07-12",
  fill = TRUE
)

weekly
```

The result uses period start and end dates rather than a hand-built week number.
That removes special January and December corrections from report scripts.

## Calculate a transparent baseline

This example uses the first four complete weeks as the reference period. Real
reports should define the reference dates according to their approved method.

```{r baseline}
reference_weeks <- sort(unique(weekly$period_start))[1:4]

baseline <- islh_surveillance_baseline(
  weekly,
  date = period_start,
  value = count,
  by = site,
  reference_periods = reference_weeks,
  method = "mean_sd",
  multiplier = 2,
  prediction_adjustment = TRUE,
  minimum_periods = 4
)

baseline
```

`reference_n` and `reference_adjustment` make the calculation auditable. The
adjustment is derived from the number of periods rather than hardcoded.
`method = "range"` returns the historical minimum and maximum, while
`method = "quantile"` uses the supplied probabilities.

A total at or above `upper_limit` is a statistical exceedance. The boundary is
inclusive on purpose: with `method = "range"` the limit is the historical
maximum, so a current total lands exactly on it regularly, and a total that has
matched the highest value ever recorded is worth reviewing. The function does
not label it an outbreak or decide whether action is required.

## Build the current seven-day table

```{r snapshot}
snapshot <- islh_surveillance_snapshot(
  daily,
  date = period_start,
  value = count,
  by = site,
  end = "2026-07-12",
  periods = 7,
  baseline = baseline,
  include_total = FALSE
)

snapshot
```

The seven displayed daily counts add to one weekly total, so the weekly
baseline above is comparable with `total`. A seven-day total must not be
compared with a baseline calculated from individual daily counts, and the
function refuses to: both tables carry the reporting period they were built
with, and mismatched durations stop with an error rather than producing a table
whose limits are seven times too small. If you display a different window,
calculate the baseline from historical totals of that same duration.

Metadata travels on the result as attributes, and subsetting a data frame drops
them. If you reshape a table between steps, name the reporting period
explicitly: pass `interval` to `islh_surveillance_baseline()`, or
`baseline_interval` to `islh_surveillance_snapshot()` for limits that came from
somewhere else entirely, such as last season's totals held in a spreadsheet.

The daily columns, `total` and baseline fields are ordinary data. Export them
with the package required by the report, or pass the result to
`islhr::islh_gt()` or `islhr::islh_flextable()`.

Only use `include_total = TRUE` when the groups are mutually exclusive. If a
record can contribute to more than one group, calculate the overall total from
the unduplicated event data instead.

## Plot the counts

`islhepi` intentionally returns data rather than imposing a figure style. Use
`islhr::islh_epi_curve()` to create the Island Health figure:

```{r plot-example, eval=requireNamespace("islhr", quietly = TRUE), fig.width=8, fig.height=4.5, fig.alt="Stacked weekly surveillance counts for Central, North and South sites."}
islhr::islh_epi_curve(
  weekly,
  date = period_start,
  count = count,
  fill = site,
  labels = "total",
  title = "Weekly surveillance counts",
  subtitle = "Complete reporting periods include zero counts",
  caption = "Source: simulated package example."
)
```

The published package website installs `islhr` and evaluates this example.
When this vignette is built on a computer without `islhr`, only the plot is
skipped; all analytical examples still run.
